Tasting Wine By the Numbers by Matthew Adkins
========================================================
  
 
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message = FALSE)
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(GGally)
library(gridExtra)
library(diptest)
library(knitr)


```

```{r echo=FALSE, Load_the_Data}
# Load the Data
winedf = read.csv('wineQualityReds.csv')
```

> This tidy data set contains 1,599 red wines with 11 variables 
on the chemical properties of the wine. 
At least 3 wine experts rated the quality of each wine, providing a rating 
between 0 (very bad) and 10 (very excellent).
Here we will explore which chemical properties influence the taste of red wines.

# Univariate Plots Section


> First let's take a look at some of the variables
and see what the values look like'

```{r echo=FALSE, Univariate_Plots}
names(winedf)
summary(winedf)
#X is not a useable variable so we will remove below
wines <- subset(winedf, select = -X)
head(wines)
```
> After removing the unneeded X variable, we can see
eleven attributes that may contribute to the taste
and quality rating of the wine.

```{r echo=FALSE, Univariate_Plots_1}
summary(wines$quality)
ggplot(wines, aes(quality))+
  geom_histogram(binwidth = 1)
```

>Here we can see that the quality ranges from
3 to 8 with a mean rating of 5.636.
Next we'll plot an analyze resiudal sugars. These
are the sugars that are left after over after red 
wines under 1 gram/liter so we shouldn't expect many
data points below or around 1.

```{r echo=FALSE, Univariate_Plots_2}
summary(wines$residual.sugar)
ggplot(wines, aes(residual.sugar))+
  geom_histogram(bins = 10)
```

>After plotting Residual Sugar within the wines we
can see that it is highly skewed to the right. Next 
we'll plot just the bottom 95% to get a better grasp
of the distribution of residual sugars'

```{r echo=FALSE, Univariate_Plots_3}
summary(subset(wines$residual.sugar,
               wines$residual.sugar <
                 quantile(wines$residual.sugar,
                          0.95)))
ggplot(wines, aes(residual.sugar))+
  geom_histogram(binwidth = .1)+
  xlim(0, quantile(wines$residual.sugar, 0.95))+
  xlab("residual.sugar (Lower 95%)")+
  geom_vline(xintercept = mean(wines$residual.sugar), color = 'blue')+
  geom_vline(xintercept = median(wines$residual.sugar), color = 'red')
```

>After elimitating the tail we can see that most wines
have a residual sugar measurement around 2.2 g/dm^3.Also
we notice that we only had one point below at 0.900 g/dm^3.
This agrees with the documentation stating the rarity of red wines
with residual sugar less than 1 gram/liter

Next we will take a look at the alcohol content of each sample
```{r echo=FALSE, Univariate_Plots_4}
summary(wines$alcohol)
ggplot(wines, aes(alcohol))+
  geom_histogram(binwidth = 0.1)+
  geom_vline(xintercept = mean(wines$alcohol), color ="blue")+
  geom_vline(xintercept = median(wines$alcohol), color='red')

```

>Again we can observe another long tail distribution with a skew to the right.
Most of the observed wines have an alcohol percentage of 9.5.

Next we'll create mutiple plots of our remaing variable and see if we see any
anomolies worth exploring'

```{r echo=FALSE, Univariate_Plots_5}
p1 <- qplot(wines$fixed.acidity)
p2 <- qplot(wines$volatile.acidity)
p3 <- qplot(wines$citric.acid)
p4 <- qplot(wines$chlorides)
p5 <- qplot(wines$free.sulfur.dioxide)
p6 <- qplot(wines$total.sulfur.dioxide)
p7 <- qplot(wines$density)
p8 <- qplot(wines$pH)
p9 <- qplot(wines$sulphates)

grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, ncol = 3)
```

>After reviewing the plots it looks like we need to adjust the chlorides graph
as it appears heavily skewed and hard to interpret in it's current raw state'.

```{r echo=FALSE, Univariate_Plots_55}
summary(wines$chlorides)
ggplot(wines, aes(x=chlorides))+
  geom_histogram()+
  scale_x_log10()+
  xlab("log10(chlorides)")
```

>Now that we transformed the chlorides graph into a normal distribution this will 
be easier for us to interpret in our analysis.

Now let's take a look at the citric acid graph. At irst take it looks like
we may have a bimodal distribution so let's explore this further

```{r echo=FALSE, Univariate_Plots_6}
qplot(wines$citric.acid, binwidth= .01)
dip.test(wines$citric.acid)
```


>After running a bimodal test we can see that our citric acid distribution 
is at least bimodal.


# Univariate Analysis



### What is the structure of your dataset?

In this data set there are 1599 observations with 12 attributes. Most of the variables are skewed
distributions. With the exception of Citric Acid (bimodal) Density (normal), and pH (normal).
It is rare to find a red wine with residual sugar under 1gram/liter and our data sample agrees with only
one sample under 1gram/liter. Also, wines with residual sugar around 45 g/l are consider sweet.
Our data set contains wines with a maximum residual sugar on 15.5 grams/liter meaning we do not have any
sweet wines within our obeservations.

### What is/are the main feature(s) of interest in your dataset?

The quality variable captures my interest the most. Quality is scored between 0-10, with 0 being the worst
and 10 being the best. Our data set shows a min of 3 and a max of 8. I am interested in trying to find what
attributes contribute to a particular wines score.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Considdering that we don't have any 'sweet' wines in our data set I believe that chlorides, pH, and
citric acid would be the contributing factors towards a wine's' overal score.

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section


>Next we will check the covariance between each of the variables.
```{r echo=FALSE, Bivariate_Plots}
round(cor(wines), 3)
ggpairs(wines)

```
>With our interests in quality alcohol, citric acid and sulphates seem to correlate
with quality of the wine the most. Let's explore this further'

```{r echo=FALSE, Bivariate_Plots_7}
ggplot(wines, aes(x=alcohol, y =quality))+
  geom_jitter(alpha = 0.25)+
  geom_smooth(method = "lm")


```

>Here we can see a positive shape to the points with our geom_smooth function pointing
up and to the right. Which suggests again that alcohol content affects quality in a positive
manner.


```{r echo=FALSE, Bivariate_Plots_8}
ggplot(wines, aes(x=alcohol, y=density))+
  geom_point(alpha = 0.25)+
  geom_smooth(method ="lm")

```

>Here we can see that the greater the alcohol the lower the density of the wine. This makes
sense given that alcohol is less dense than water. Thus more alcohol would make the density fall.


```{r echo=FALSE, Bivariate_Plots_9}
ggplot(wines, aes(x= residual.sugar, y=quality))+
  xlim(0, quantile(wines$residual.sugar,0.95))+
  geom_jitter(alpha=0.15)+
  xlab("residual sugar (lower 95%)")

```

> As seen in our covariance residual sugar has almost no influence on the quality of our sample of wines.
Perhaps if our study was targeted towards sweet wines we would see a correlation.

```{r echo=FALSE, Bivariate_Plots_10}
ggplot(wines, aes(x=pH, y=quality))+
  geom_jitter(aplha=0.25)+
  geom_smooth(method ="lm")


```

> Here it becomes apparent that wines with higher pH values(more basic) are generally rated lower than
wines with lower pH values. There does seem to be a range between 3.0 and 3.5 where most of our data set falls.

```{r echo=FALSE, Bivariate_Plots_11}
ggplot(wines, aes(x=density, y=fixed.acidity))+
  geom_jitter(alpha=0.25)+
  geom_smooth(method = "lm")


```

> Here we can see that fixed acidity has a highly positive correlation with density. 

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I found it interesting that alcohol content had such an influence on quality score. I initially expected
citric acid to be the highest contributor has it adds a freshness and flavor to the wine. We also observed that 
residual sugar had almost no influence on our wine scores. However, this may have been different if we were studying 
sweet wines.


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
I found it really interesting that fixed acidity was more correlated to density than alcohol. Though alcohol negatively contributes
to the density of the wine, fixed acidity positively contributed to density. This makes me curious as to what chemicals add to the acidity
and how heavy those molecules are.

### What was the strongest relationship you found?

It may be no suprise but attributes that contribute to the pH value of the wine are highly correlated with eachother. I say no surprise
because these factors are fixed.acidity, citric.acid, and density. The first two both directly contribute to pH but again the most surprising
relationship was fixed.acidity and density.


# Multivariate Plots Section


> Next we'll' use a multivariate approach to help find prdeictive variable of quality. We'll'
start plotting alcohol content vs quality and color in each point according to it's citric' acid
rating. Blue dots will represent lowest quantity of citric acid while red will be the highest. A gradient
function will be used to show varying degrees of citric acid present.


```{r echo=FALSE, Multivariate_Plots}
ggplot(wines, aes(x= alcohol, y= quality, color =citric.acid))+
  geom_jitter()+
  scale_color_gradient(low = 'blue', high = 'red')

```
 >Here a small pattern arises with higher concentrations of ctric acid in the upper right corner.


```{r echo=FALSE, Multivariate_Plots_12}
ggplot(wines, aes(x= alcohol, y= density, color = fixed.acidity))+
  geom_jitter()+
  scale_color_gradient(low = 'blue', high = 'red')


```

> This plot shows how alcohol content and fixed acidity have a drastic effect on the density of the wine. It appears that these
two attributes counter-balance eachother. Alcohol lowers the denisty while more fixed acidity will raise the density of the wine. 



```{r echo=FALSE, Multivariate_Plots_13}
ggplot(wines, aes(x= alcohol, y= citric.acid, color = factor(quality)))+
  geom_jitter()+
  scale_color_brewer()

```


> Now when we analyze citric acid and alcohol we can see that the quality score tends towards a positive when both are present. 

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

It seems that a possibly reliable predictor is the presence of both citric acid and alcohol content. Citric acid adds freshness and flavor
to the wine but this alone was not enough to raise a wine's' quality score. However, if citric acid is paired with alcohol content
you could potentially produce a higher scoring wine.

### Were there any interesting or surprising interactions between features?

An interesting relationship was discovered between fixed.acidity and alcohol. Alcohol lowers density while fixed acidity seems to contribute
to a denser wine. Given that alcohol is such a high contributor to quality score it suggests that fixed.acid would lower the wine's'
over all score.



------

# Final Plots and Summary



### Plot One
```{r echo=FALSE, Plot_One}
ggplot(wines, aes(alcohol))+
  geom_histogram(binwidth = .1)+
  geom_vline(xintercept = median(wines$alcohol), color = 'blue')+
  geom_vline(xintercept = mean(wines$alcohol), color = 'red')+
  ggtitle("Distribution of Alcohol Percentages")+
  xlab("Alcohol (%)")+
  ylab("Number of Wines")+
  annotate('text',
           x = median(wines$alcohol) - 0.35,
           y = 120,
           label = paste('median\n(', median(wines$alcohol), ')', sep = ''),
           color = 'blue')+
  annotate('text',
           x = mean(wines$alcohol) + 0.35,
           y = 120,
           label = paste('mean\n(', round(mean(wines$alcohol), 2), ')', sep = ''),
           color = 'red')
  

```

### Description One

Here we can see that the distribution of our observed wines from the data set.
This distribution is a right-skewed distribution with very few wines exceeding 12%
alcohol.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(wines, aes(x=alcohol, y=quality))+
  geom_jitter(alpha = .1, height =0.5, width =0.025)+
  geom_smooth(method = "lm")+
  ggtitle("Quality vs Alcohol COntent")+
  xlab("Alcohol (%)")+
  ylab("Quality (0-10)")
  
```

### Description Two

Here we are able to observe the positve effect alcohol content has on the quality rating of an individual wine. 
Most wines are below 10% but those wines that can achieve an alcohol rating tend to have a higher rating.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(wines, aes(x=alcohol, y = citric.acid, color=factor(quality)))+
  geom_jitter()+
  scale_color_brewer(name = "Quality")+
  ggtitle("Quality by Citric Acid and Alcohol")+
  xlab("Alcohol (%)")+
  ylab("Citric Acid (g/L)")

```

### Description Three

Citric acid adds to the wine's' flavor and tase of freshness. However, though it directly determines the wines flavor
it is not enough to raise a wine's' quality score. Instead it needs a helping hand from the wine's' alcohol content 
to help provide a better scoring wine.

------
  
# Reflection
  
  This project data set consisted of 1599 observations and eleven attributes that could effect overall quality of the
wine. Surprisingly, certain factors that should have a direct effect on taste, and therefore quality, were not always
enough to determine a wine's' quality rating. Factors such as citric acid directly determine a wine's' flavor however
it did not change the wine's' quality rating. Other factors such as alcohol content very strongly determined a wine's'
rating. In the case of alcohol content more is better.

We were also able to see that fixed acidity correlated with density. This implies that the process that raises fixed
acidity also creates a denser wine. We are unable to determine if this effects the flavor of the wine but it does counter-act
the density lowering factor of the alcohol content in the wine. 

A final thought is that we found that our data set consisted of rather mild red wines and did not contain any of the "sweet"
varieties, wine with 45 g/l or more. Thus in our data set residual sugar was not a determining factor when analyzing our wines. 
In futures it would be interesting to study residual sugar in wines that are meant to be sweeter. Would more residual sugar
lend to better quality wines? Or does this variable only become useful in classifying the kind of wine in the data set?
  
  